Comparing and combining network-guided biomarker discovery to discover genetic susceptibility mechanisms to breast cancer on the GENESIS study

### 2021-01-26 manuscript/plos/review_2.tex
added chloe's comments

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Genome-wide association studies (GWAS) explore the genetic causes of complex diseases. However, they ignore the biological context of the genetic variants and genes under study. To address this shortcoming, one can seek susceptibility genes that are functionally related, employing networks to integrate such functional information. Many such network methods exist, each stemming from different mathematical frameworks, pre-processing steps, and assumptions about the susceptibility mechanism. Unsurprisingly, their results are also disparate. To sort through this diversity, we selected six network methods and applied them to GENESIS, a GWAS on French women with non-BRCA familial breast cancer. First, we verified that network methods recovered more interesting and interpretable genes than a standard GWAS. Nevertheless, as expected, these results were notably different. We addressed this heterogeneity studying their overlap, computing what we called the \emph{consensus}. The key gene in this consensus solution was \emph{COPS5}, a gene related to multiple cancer hallmarks. Another issue of network methods was their instability, selecting very different genes on different subsamples of GENESIS. Therefore, we proposed a \emph{stable consensus} solution formed by the 68 genes most consistently selected across multiple subsamples. Importantly, this solution was also enriched in genes associated with breast cancer susceptibility (\emph{BLM}, \emph{CASP8}, \emph{CASP10}, \emph{DNAJC1}, \emph{FGFR2}, \emph{MRPS30}, and \emph{SLC4A7}, P\=/value~=~$3 \times 10^{-4}$). The most connected gene was \emph{CUL3}, a regulator of several genes linked to cancer progression. Lastly, we evaluated the biases of each method and the impact of the parameters on the outcome. In general, network methods preferred highly connected genes, even after random rewirings that stripped the connections of any biological meaning. In conclusion, we presented the advantages of network-guided GWAS, characterized their shortcomings, and provided strategies to address them. To compute the consensus networks, implementations of all six methods are available at \url{https://github.com/hclimente/gwas-tools}.

>>>>

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Genome-wide association studies (GWAS) explore the genetic causes of complex diseases. However, they ignore the biological context of the genetic variants and genes under study. To address this shortcoming, one can use biological networks, which model functional relationships, to search for functionally related susceptibility loci. Many such network methods exist, each arising from different mathematical frameworks, pre-processing steps, and assumptions about the susceptibility mechanism. Unsurprisingly, this results in disparate solutions. To explore how to exploit these heterogeneous approaches, we selected six network methods and applied them to GENESIS, a GWAS on French women with non-BRCA familial breast cancer. First, we verified that network methods recovered more interpretable genes than a standard GWAS. We addressed the heterogeneity of their solutions by studying their overlap, computing what we called the \emph{consensus}. The key gene in this consensus solution was \emph{COPS5}, a gene related to multiple cancer hallmarks. Another issue we observed was that network methods were unstable, selecting very different genes on different subsamples of GENESIS. Therefore, we proposed a \emph{stable consensus} solution formed by the 68 genes most consistently selected across multiple subsamples. This solution was also enriched in genes associated with breast cancer susceptibility (\emph{BLM}, \emph{CASP8}, \emph{CASP10}, \emph{DNAJC1}, \emph{FGFR2}, \emph{MRPS30}, and \emph{SLC4A7}, P\=/value~=~$3 \times 10^{-4}$). The most connected gene was \emph{CUL3}, a regulator of several genes linked to cancer progression. Lastly, we evaluated the biases of each method and the impact of their parameters on the outcome. In general, network methods preferred highly connected genes, even after random rewirings that stripped the connections of any biological meaning. In conclusion, we presented the advantages of network-guided GWAS, characterized their shortcomings, and provided strategies to address them. To compute the consensus networks, implementations of all six methods are available at \url{https://github.com/hclimente/gwas-tools}.
